Geometrization of quasicharacters of algebraic tori over local fields

### 2014-01-22 geometrization.tex
Fixed some references; added a reference to my paper with Masoud.

\begin{abstract}
For any non-archimedean local field $K$
  and any algebraic torus $T$ over $K$, we use the
  function--sheaf dictionary for smooth group schemes locally of finite type over finite fields
  to geometrize quasicharacters of $T(K)$
  as quasicharacter sheaves on the Greenberg transform of the Néron
  model of $T$.
 As an application, we categorify a transfer of certain characters between characteristic $p$ tori and characteristic $0$ tori.
  \end{abstract}

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We would like to thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable. 
We also thank Alessandra Bertrapelle and Cristian Gonz\'ales-Avil\'es for disabusing us of a misapprehension concerning the Greenberg realization functor and for pointing us to a proof of Lemma~\ref{lemma:wrK}.

@@

  Finally, we show that $\trFrob{T,n}$ is functorial.
The first part of Proposition~\ref{prop:pullback} shows that $F_1$ is a functor,
while the second part shows that trace of Frobenius is a natural transformation
$t_{ - ,n} : F_1 \to F_2$. When further combined with Proposition~\ref{prop:product},
we see that $F_1$ is an additive functor and $t_{ - ,n}: F_1 \to F_2$ is a
natural transformation between additive functors.
\end{proof}

@@

For any  ordered pair of positive integers $n< m$ define 
$f_{n\leq m} : \GN{T}_{m} \to \GN{T}_{n}$ 
by $f_{n\leq m}\ceq f_n \circ f_{n+1} \circ \cdots \circ f_{m-1}$; 
also set $f_{n\leq n} \ceq \id_{\GN{T}_n}$.
Note that $f_{n\leq m}:  \GN{T}_{m} \to \GN{T}_{n}$ is a surjective morphism 
of group schemes over $\Fq$.
Let 
\[
f_{n\leq m}^* : \QC(\GN{T}_n) \to \QC(\GN{T}_{m})
\]
be the functor produced by applying Proposition~\ref{prop:pullback} to $f_{n\leq m}$.
We now define a category which allows us to extend the notion 
of quasicharacter sheaf to the pro-group scheme $\GN{T}$.

@@

\begin{proof}
We may define full and faithful functors $F_n : \QC(\GN{T}_n) \to \QC(\GN{T})$
for each $n$ by $F_n(\qcs{L}) \ceq [\qcs{L}_n]$ and $F_n(\alpha_n) \ceq [\alpha_n]$.
For positive integers $n \le m$ these functors fit together into a diagram
\[
\begin{tikzcd}
\ & & & \QC(\GN{T})\\
\arrow{urrr}{F_n} \QC(\GN{T}_n) \arrow[swap]{rr}{f_{n\leq m}^*} && \QC(\GN{T}_m) \arrow[swap]{ur}{F_m}. & 
\end{tikzcd}
\]
Passing to isomorphism classes of objects and invoking the 
universal property of direct limits gives a group isomorphism
$\theta: \varinjlim_{n\in \ZZ_{>0}} \QCiso{\GN{T}_n} \to \QCiso{\GN{T}}$
with inverse $[\qcs{L}]_{/\text{iso}} \mapsto [\qcs{L}_{/\text{iso}}]$.
Now use Theorem~\ref{thm:application}, Proposition~\ref{prop:pullback} 
and the fact that  $\Hom_{\text{qc}}(T(K),\EEx) = \varinjlim_{n\in \ZZ_{>0}} \GN{T}_n(k)^*$, 
where this direct limit is taken with respect to the family of group homomorphisms 
dual to $f_{n} : \GN{T}_{n+1}(\Fq) \to \GN{T}_n(\Fq)$.
\end{proof}

@@

\begin{proof}
Adapting the argument in \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 7.6, Prop 6} to locally finite type Néron models,
one may show that $\Res_{\OKp/\OK}(\TTp)$ is a Néron model for $\Res_{K'/K}(T')$.
By the Néron mapping property, the canonical closed immersion
$T\hookrightarrow \Res_{K'/K}(T')$ extends uniquely to a morphism
$\underline{T}\to \Res_{\OKp/\OK}(\TTp)$ of smooth $\OK$-group schemes.
Applying the Greenberg transform $\Gr_{n}$ to this morphism
and using Lemma~\ref{lemma:wrK} defines the morphism of $\Fq$-group schemes
\begin{equation}\label{men}
\GN{T}_n \to \Res_{k'/k} \GN{T}'_{en}.
\end{equation}
Passing to the pro-group schemes in Definition~\ref{def:objGN} 
the family of morphisms \eqref{men} determines a morphism of pro-group schemes
\[
\GN{T} \to \Res_{k'/k} \GN{T}'.
\]
Since $\left(\Res_{k'/k} \GN{T}'_{en} \right)(\Fq) = \left(\GN{T}'_{en}\right)(k')$,
the pullback functor $\QC(\Res_{k'/k} \GN{T}'_{en})\to \QC(\GN{T}_n)$
obtained by applying Proposition~\ref{prop:pullback} to \eqref{men} gives
\[
\Hom_{< en}(T'(K'),\EEx) \to \Hom_{< n}(T(K),\EEx).
\]
Passing to inverse limits defines
\[
\Hom_{\text{qc}}(T(K'),\EEx) \to \Hom_{\text{qc}}(T(K),\EEx).
\]
Argue as in Proposition~\ref{prop:csbe} to see that this is indeed restriction of characters.
\end{proof}

@@

We have given a functorial relationship between quasicharacter sheaves 
on the Greenberg transform of the Néron model of $T$ 
and quasicharacters of $T(K)$.
When coupled with the reciprocity map $\text{rec}_T $ for $T$, 
which enjoys the same functorial properties, 
it follows that quasicharacter sheaves for $T$ 
functorially correspond to Langlands parameters for $T$.
We suspect that it is possible to extract Langlands parameters 
from quasicharacter sheaves directly and that the reciprocity map 
can then be viewed as a matching between quasicharacters and Langlands parameters coming
 from the same quasicharacter sheaf. 
 \[
\begin{tikzcd}
\ & \arrow[swap]{dl}{\trFrob{G}} \QCiso{\GN{T}} \arrow[dashed]{dr} & \\
\Hom_{\text{qc}}(T(K),\EEx) \arrow{rr}{\text{rec}_T} && H^1(K,\hat{T}_\ell)
\end{tikzcd}
\]
 Results in this paper provide tools for reduction of the problem to the case $T=\Gm{K}$. 
 In that case, our suspicions are all but confirmed 
 by the class field theory of Serre-Hazewinkel
 \citelist{\cite{serre:60a} \cite{serre:61a} \cite{demazure-gabriel:GroupesAlgebriques}*{Appendice}},
 especially as revisited in \cite{suzuki-yoshida:12a}*{\S 2}. 
 What remains to be done is a careful comparison of 
 quasicharacter sheaves for $\Gm{K}$ and the local systems on
  the fpqc-site of the pro-quasi-algebraic varieties appearing in \cite{suzuki-yoshida:12a}.

>>>>

\begin{abstract}
For any non-archimedean local field $K$
  and any algebraic torus $T$ over $K$, we use the
  function--sheaf dictionary for smooth group schemes locally of finite type over finite fields
  to geometrize quasicharacters of $T(K)$
  as quasicharacter sheaves on the Greenberg transform of the Néron
  model of $T$.
 As an application, we find a categorical transfer of certain characters between characteristic $p$ tori and characteristic $0$ tori.
  \end{abstract}

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
%We would like to thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of quasicharacter sheaves; their knowledge and help have been invaluable. 
We thank Alessandra Bertrapelle and Cristian Gonz\'ales-Avil\'es for disabusing us of a misapprehension concerning the Greenberg realization functor and for pointing us to a proof of Lemma~\ref{lemma:wrK}.

@@

  Finally, we show that $\trFrob{T,n}$ is functorial.
The first part of \cite{cunningham-roe:dictionary}*{Prop. 4.1} shows that $F_1$ is a functor,
while the second part shows that trace of Frobenius is a natural transformation
$t_{ - ,n} : F_1 \to F_2$. When further combined with \cite{cunningham-roe:dictionary}*{Prop. 4.2},
we see that $F_1$ is an additive functor and $t_{ - ,n}: F_1 \to F_2$ is a
natural transformation between additive functors.
\end{proof}

@@

For any  ordered pair of positive integers $n< m$ define 
$f_{n\leq m} : \GN{T}_{m} \to \GN{T}_{n}$ 
by $f_{n\leq m}\ceq f_n \circ f_{n+1} \circ \cdots \circ f_{m-1}$; 
also set $f_{n\leq n} \ceq \id_{\GN{T}_n}$.
Note that $f_{n\leq m}:  \GN{T}_{m} \to \GN{T}_{n}$ is a surjective morphism 
of group schemes over $\Fq$.
Let 
\[
f_{n\leq m}^* : \QC(\GN{T}_n) \to \QC(\GN{T}_{m})
\]
be the functor produced by applying \cite{cunningham-roe:dictionary}*{Prop. 4.1} to $f_{n\leq m}$.
We now define a category which allows us to extend the notion 
of quasicharacter sheaf to the pro-group scheme $\GN{T}$.

@@

\begin{proof}
We may define full and faithful functors $F_n : \QC(\GN{T}_n) \to \QC(\GN{T})$
for each $n$ by $F_n(\qcs{L}) \ceq [\qcs{L}_n]$ and $F_n(\alpha_n) \ceq [\alpha_n]$.
For positive integers $n \le m$ these functors fit together into a diagram
\[
\begin{tikzcd}
\ & & & \QC(\GN{T})\\
\arrow{urrr}{F_n} \QC(\GN{T}_n) \arrow[swap]{rr}{f_{n\leq m}^*} && \QC(\GN{T}_m) \arrow[swap]{ur}{F_m}. & 
\end{tikzcd}
\]
Passing to isomorphism classes of objects and invoking the 
universal property of direct limits gives a group isomorphism
$\theta: \varinjlim_{n\in \ZZ_{>0}} \QCiso{\GN{T}_n} \to \QCiso{\GN{T}}$
with inverse $[\qcs{L}]_{/\text{iso}} \mapsto [\qcs{L}_{/\text{iso}}]$.
Now use Theorem~\ref{thm:application}, \cite{cunningham-roe:dictionary}*{Prop. 4.1} 
and the fact that  $\Hom_{\text{qc}}(T(K),\EEx) = \varinjlim_{n\in \ZZ_{>0}} \GN{T}_n(k)^*$, 
where this direct limit is taken with respect to the family of group homomorphisms 
dual to $f_{n} : \GN{T}_{n+1}(\Fq) \to \GN{T}_n(\Fq)$.
\end{proof}

@@

\begin{proof}
Adapting the argument in \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 7.6, Prop 6} to locally finite type Néron models,
one may show that $\Res_{\OKp/\OK}(\TTp)$ is a Néron model for $\Res_{K'/K}(T')$.
By the Néron mapping property, the canonical closed immersion
$T\hookrightarrow \Res_{K'/K}(T')$ extends uniquely to a morphism
$\underline{T}\to \Res_{\OKp/\OK}(\TTp)$ of smooth $\OK$-group schemes.
Applying the Greenberg transform $\Gr_{n}$ to this morphism
and using Lemma~\ref{lemma:wrK} defines the morphism of $\Fq$-group schemes
\begin{equation}\label{men}
\GN{T}_n \to \Res_{k'/k} \GN{T}'_{en}.
\end{equation}
Passing to the pro-group schemes in Definition~\ref{def:objGN} 
the family of morphisms \eqref{men} determines a morphism of pro-group schemes
\[
\GN{T} \to \Res_{k'/k} \GN{T}'.
\]
Since $\left(\Res_{k'/k} \GN{T}'_{en} \right)(\Fq) = \left(\GN{T}'_{en}\right)(k')$,
the pullback functor $\QC(\Res_{k'/k} \GN{T}'_{en})\to \QC(\GN{T}_n)$
obtained by applying \cite{cunningham-roe:dictionary}*{Prop. 4.1} to \eqref{men} gives
\[
\Hom_{< en}(T'(K'),\EEx) \to \Hom_{< n}(T(K),\EEx).
\]
Passing to inverse limits defines
\[
\Hom_{\text{qc}}(T(K'),\EEx) \to \Hom_{\text{qc}}(T(K),\EEx).
\]
Argue as in \cite{cunningham-roe:dictionary}*{Prop. 5.1} to see that this is indeed restriction of characters.
\end{proof}

@@

We have given a functorial relationship between quasicharacter sheaves 
on the Greenberg transform of the Néron model of $T$ 
and quasicharacters of $T(K)$.
When coupled with the reciprocity map $\text{rec}_T $ for $T$, 
which enjoys the same functorial properties, 
it follows that quasicharacter sheaves for $T$ 
functorially correspond to Langlands parameters for $T$.
We suspect that it is possible to extract Langlands parameters 
from quasicharacter sheaves directly and that the reciprocity map 
can then be viewed as a matching between quasicharacters and Langlands parameters coming
 from the same quasicharacter sheaf. 
 \[
\begin{tikzcd}
\ & \arrow[swap]{dl}{\trFrob{G}} \QCiso{\GN{T}} \arrow[dashed]{dr} & \\
\Hom_{\text{qc}}(T(K),\EEx) \arrow{rr}{\text{rec}_T} && H^1(K,\hat{T}_\ell)
\end{tikzcd}
\]
 Results in this paper provide tools for reduction of the problem to the case $T=\Gm{K}$. 
 In that case, our suspicions are all but confirmed 
 by the class field theory of Serre-Hazewinkel
 \citelist{\cite{serre:60a} \cite{serre:61a} \cite{demazure-gabriel:GroupesAlgebriques}*{Appendice}},
 especially as revisited in \cite{suzuki-yoshida:12a}*{\S 2}. 
 What remains to be done is a careful comparison of 
 quasicharacter sheaves for $\Gm{K}$ and the local systems on
  the fpqc-site of the pro-quasi-algebraic varieties appearing in \cite{suzuki-yoshida:12a}. 
In this regard, a comparison between the results in this paper and those in \cite{cunningham-kamgarpour:geometrization} might also be illuminating.
